
TB knitr lecture notes
======================

## Global options

Could set global options:
```{r setoptions, echo=FALSE}
# opts_chunk$set(echo = FALSE, results = "hide")
```

## Other common options for a given chunk

don't forget the two spaces to carriage return  

results = "asis" or "hide"  
fig.height = numeric  
fig.width = numeric  
echo = TRUE or FALSE  

cache=TRUE (for a given chunk)
    * it will have to run it the first time
    * if the data or code change, it will re-run

## Introduction

if not using Rstudio:
```{r firstchunk}
#library(knitr)
#setwd(<working directory>)
#knit2html("blah.Rmd")
#browseURL("blah.html")
```

Knitr takes the .Rmd and makes a markdown file - see lecture notes

```
numerical output goes in a box like this in the markdown file
```


```{r scatterplot, echo=FALSE, fig.height=4}
plot(cars)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.

```{r computetime, echo=FALSE}
time <- format(Sys.time(), "%a %b %d %X %Y")
rand <- rnorm(1)
```

The current time is `r time`.  My favorite number is `r rand`. 


## Incorporating Graphics

Let's first simulate some data
```{r simulate data, echo=TRUE}
x <- rnorm(100); y <- x + rnorm(100, sd = 0.5)
```
Here is a scatterplot of those data
```{r scatterplot2, fig.height=5}
par(mar = c(5,4,1,1), las=1)
plot(x,y, main = "My Simulated Data")
```

## Making Tables with xtable

```{r fitmodel}
library(datasets)
data(airquality)
fit <- lm(Ozone ~ Wind + Temp + Solar.R, data = airquality)
```
Here is a table of regression coefficients.
```{r xtable, results="asis"}
library(xtable)
xt <- xtable(summary(fit))
print(xt, type="html")
```

## Trying out stargazer instead - not working yet

```{r stargazer}
library(stargazer)
stargazer(fit, title="Table") ### TB this gets you Latex

stargazer(fit, type="html", title="Table") ### this gets you raw html
# print(st)
```

## Caching computations

