---
output: html_document
---
## Notes from B Caffo week 3 lectures


### T Confidence intervals

* In the previous, we discussed creating a confidence interval using the CLT
    - They took the form $Est \pm ZQ \times SE_{Est}$
*In this lecture, we discuss some methods for small samples, notably Gosset's $t$ distribution and $t$ confidence intervals
    - They are of the form $Est \pm TQ \times SE_{Est}$
* These are some of the handiest of intervals
* If you want a rule between whether to use a $t$ interval or normal interval, just always use the $t$ interval
* We'll cover the one and two group versions

### Gosset's $t$ distribution

* Invented by William Gosset (under the pseudonym "Student") in 1908
* Has thicker tails than the normal
* Is indexed by a degrees of freedom; gets more like a standard normal as df gets larger 
* It assumes that the underlying data are iid Gaussian with the result that $$ \frac{\bar X - \mu}{S/\sqrt{n}} $$ follows Gosset's $t$ distribution with $n-1$ degrees of freedom
    
* (If we replaced $s$ by $\sigma$ the statistic would be exactly standard normal)
* Interval is $\bar X \pm t_{n-1} S/\sqrt{n}$ where $t_{n-1}$ is the relevant quantile

### Code for manipulate

```{r tplot}
library(ggplot2); library(manipulate)
k <- 1000
xvals <- seq(-5, 5, length = k)
myplot <- function(df){
  d <- data.frame(y = c(dnorm(xvals), dt(xvals, df)),
                  x = xvals,
                  dist = factor(rep(c("Normal", "T"), c(k,k))))
  g <- ggplot(d, aes(x = x, y = y)) 
  g <- g + geom_line(size = 2, aes(colour = dist))
  g
}
manipulate(myplot(mu), mu = slider(1, 20, step = 1))  
```


### Easier to see

```{r quantilesofthet}
pvals <- seq(.5, .99, by = .01)
myplot2 <- function(df){
  d <- data.frame(n= qnorm(pvals),t=qt(pvals, df),
                  p = pvals)
  g <- ggplot(d, aes(x= n, y = t))
  g <- g + geom_abline(size = 2, col = "lightblue")
  g <- g + geom_line(size = 2, col = "black")
  g <- g + geom_vline(xintercept = qnorm(0.975))
  g <- g + geom_hline(yintercept = qt(0.975, df))
  g
}
manipulate(myplot2(df), df = slider(1, 20, step = 1))
```

### Notes about the $t$ interval

* The $t$ interval technically assumes that the data are iid normal, though it is robust to this assumption
* It works well whenever the distribution of the data is roughly symmetric and mound shaped
* Paired observations are often analyzed using the $t$ interval by taking differences
* For large degrees of freedom, $t$ quantiles become the same as standard normal quantiles; therefore this interval converges to the same interval as the CLT yielded
    
* For skewed distributions, the spirit of the $t$ interval assumptions are violated
    - Also, for skewed distributions, it doesn't make a lot of sense to center the interval at the mean
    - In this case, consider taking logs or using a different summary like the median
* For highly discrete data, like binary, other intervals are available

### Sleep data

In R typing data(sleep) brings up the sleep data originally analyzed in Gosset's Biometrika paper, which shows the increase in hours for 10 patients on two soporific drugs. R treats the data as two groups rather than paired.
The data

`r data(sleep)`
`r head(sleep)`

### Plotting the data

```{r sleepplot}
library(ggplot2)
g <- ggplot(sleep, aes(x = group, y = extra, group = factor(ID)))
g <- g + geom_line(size = 1, aes(colour = ID)) + geom_point(size =10, pch = 21, fill = "salmon", alpha = .5)
g
```

### Results

```{r sleepresults}
g1 <- sleep$extra[1 : 10]; g2 <- sleep$extra[11 : 20]
difference <- g2 - g1
mn <- mean(difference); s <- sd(difference); n <- 10

mn + c(-1, 1) * qt(.975, n-1) * s / sqrt(n)
t.test(difference)
t.test(g2, g1, paired = TRUE)
t.test(extra ~ I(relevel(group, 2)), paired = TRUE, data = sleep)
```
