---
title: "Severe Weather Event Analysis"
author: "Tell Bennett"
date: "Monday, April 20, 2015"
output: html_document
---
    Title: Your document should have a title that briefly summarizes your data analysis

    Synopsis: Immediately after the title, there should be a synopsis which describes and summarizes your analysis in at most 10 complete sentences.

### Synopsis

10 sentences or less.



### Data Processing

```{r options, echo = FALSE}
options(dplyr.max_print = 10)
options(dplyr.min_print =  5)
```

1) Loading packages

```{r loadpackages, echo = FALSE}
library(dplyr)
library(lubridate)
library(ggplot2)
```

2) Loading the raw data

```{r dataload}
url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"


## don't forget time
```

3) Reference documentation
Put in a hyperlink to https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf

FAQ location 

### Results


When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document. You can embed an R code chunk like this:

```{r}
summary(cars)
```

You can also embed plots, for example:

```{r, echo=FALSE}
plot(cars)
```

